---
title: "discrimability"
author: "Greg Kiar"
date: "January 13, 2017"
output:
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 11
    highlight: pygments
    keep_md: yes
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
---

# Discriminability in a Nutshell

We are going to generate data from two classes (circles and squares), and demonstrate the
strength of discriminability through evaluating different sampling or processing strategies.
We construct our two class distributions, sample points from them, reconstruct the object,
compute the discriminability of the processing. Discriminability is: computing distance between
all pairs of objects, rank the distances, compute the histogram of same-class ranks, and compute
the mean rank of the distribution.

Before getting started, we will define a few helpful functions and variables which will streamline
this process.

```{r load, warning=FALSE}
library(abind)
library(rmarkdown)
library(knitr)
library(RColorBrewer)
library(gplots)
```

```{r knitr-setup, include=FALSE, results='asis'}
### The following options and figure numbering functions
### were setup by Youngser Park
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE)
dep_auto() # figure out dependencies automatically
opts_chunk$set(cache=FALSE,echo=TRUE,warning=FALSE,message=FALSE,
               comment="#",
               fig.path='./Figures/',
               dpi=227,dev=c('png','pdf'))

opts_knit$set(aliases=c(h='fig.height', w='fig.width', 
                        cap='fig.cap', scap='fig.scap'))
opts_knit$set(eval.after = c('fig.cap','fig.scap'))
knit_hooks$set(document = function(x) {
  gsub('(\\\\end\\{knitrout\\}[\n]+)', '\\1\\\\noindent ', x)
})
#opts_knit$set(animation.fun = hook_scianimator)

knit_hooks$set(plot = function(x, options) {
       paste('<figure><img src="',
             opts_knit$get('base.url'), paste(x, collapse = '.'),
             '"><figcaption>', options$fig.cap, '</figcaption></figure>',
             sep = '')
 })
 
 fn = local({
   i = 0
   function(x) {
     i <<- i + 1
     paste('Figure ', i, ': ', x, sep = '')
   }
 })

 fig <- local({
    i <- 0
    ref <- list()
    list(
        cap=function(refName, text) {
            i <<- i + 1
            ref[[refName]] <<- i
            paste("<b>Figure ", i, ": ", text, "</b><br><br>", sep="")
        },
        ref=function(refName) {
            ref[[refName]]
        })
})
```

```{r setup}
lweight <- 4
len <- 500
nsamples <- 40
nsims <- 3
noise1 <- 0.15
tsize <- 4
darkcol <- '#28282e'
lightcol <- '#98929e'

circle <- function(){
  theta <- seq(from = 0, to = 2*pi - pi/len, by = 2*pi/len)
  x <- cos(theta)
  y <- sin(theta)
  circ <- list(x, y)
}

square <- function(){
  range <- seq(from = -1, to = 1, by = 2/(len/4))
  static <- rep(1, each = len/4)
  x <- c(range, static, -range, -static) 
  y <- c(static, -range, -static, range)
  sqr <- list(x, y)
}

shape_plot <- function(xs, ys, typ){
  size <- lweight
  if (typ == 'p'){
    size <- size/3
  }
  plot(xs, ys, type=typ, axes=FALSE, xlab='', ylab='', pch=20, asp=1, lwd=size, col=darkcol)
  if (typ == 'l'){
    polygon(x=xs, y=ys, col=lightcol)
  }
}

sample1 <- function(data, n){
  pts <- sort(floor(runif(n, min=1, max=len)))
  xs <- data[[1]]
  x <- xs[pts] + noise1*runif(n, min=-1, max=1)
  x <- c(x, x[1])
  ys <- data[[2]]
  y <- ys[pts] + noise1*runif(n, min=-1, max=1)
  y <- c(y, y[1])
  samp <- list(x, y)
}

distances <- function(data){
  diff <- matrix(data=NA, nrow=2*nsims, ncol=2*nsims)
  for (i in 1:(2*nsims)){
    for (j in 1:(2*nsims)){
      diff[i,j] <- norm(data[,,i] - data[,,j], '2')
    }
  }
  cols <- colorRampPalette(c("green", "yellow", "red"))(n = 299)
  heatmap.2(diff, cellnote=round(diff,1), Colv=NA, Rowv=NA, trace="none",
            density.info="none", notecol="black", col=cols, cexRow=tsize,
            cexCol=tsize, notecex=tsize, labRow="", labCol="", key=FALSE,
            lwid=c(.5,6), lhei=c(.5,6), margins=c(4.5,4.5))

}

simulate_shape <- function(fn){
  layout(matrix(c(1,2,3, 1,4,5, 1,6,7), 3, 3, byrow = TRUE))
  par(xpd=NA)
  shape <- fn()
  shape_plot(shape[[1]], shape[[2]], 'l')
  
  arrows(x0=1.2, x1=2.2, y0=1, y1=1.5, length=0.1, lwd = lweight)
  arrows(x0=1.3, x1=2.2, y0=0, y1=0, length=0.1, lwd = lweight)
  arrows(x0=1.2, x1=2.2, y0=-1, y1=-1.5, length=0.1, lwd = lweight)
  text(x=1.5, y=1.5, expression('s'[1]^'a'), cex=tsize)
  text(x=1.7, y=0.3, expression('s'[2]^'a'), cex=tsize)
  text(x=1.5, y=-1.5, expression('s'[3]^'a'), cex=tsize)

  arrows(x0=3.5, x1=4.8, y0=1.5, y1=1.5, length=0.1, lwd=lweight)
  arrows(x0=3.5, x1=4.8, y0=0, y1=0, length=0.1, lwd=lweight)
  arrows(x0=3.5, x1=4.8, y0=-1.5, y1=-1.5, length=0.1, lwd=lweight)

  text(x=4.15, y=1.8, bquote(p[.(1)]^'a'), cex=tsize)
  text(x=4.15, y=0.3, bquote(p[.(2)]^'a'), cex=tsize)
  text(x=4.15, y=-1.2, bquote(p[.(3)]^'a'), cex=tsize)
  
  # label <- rep(as.character(substitute(fn)), nsims)
  label <- c()
  data <- array(data=NA, c(2, nsamples, nsims))
  dim(data)
  for (i in 1:nsims){
    samp <- sample1(shape, nsamples)
    shape_plot(samp[[1]], samp[[2]], 'p')
    shape_plot(samp[[1]], samp[[2]], 'l')
    data[1, ,i] <- samp[[1]][1:nsamples]
    data[2, ,i] <- samp[[2]][1:nsamples]
    label <- c(label, as.character(substitute(fn)))
  }
  values <- list(label, data)
}
```

## Sampling shapes

We will start by sampling data from the circle distribution and reconstructing the shapes
from each set of samples.

```{r circle_sampling, fig.align='center'}
# pdf("circle_plot.pdf", width=20, height=10)
vals <- simulate_shape(circle)
# dev.off()
labels <- vals[[1]]
data <- vals[[2]]
```

The same process is then repeated for the square distribution.
```{r square_sampling, fig.align='center'}
vals <- simulate_shape(square)
labels <- c(labels, vals[[1]])
data <- abind(data, vals[[2]])
```

Now we have a data matrix which contains `x` and `y` sampled values, and a list of labels
for each sample identifying whether it belongs to the "circle" or "square" distribution.
We can then compute the distance between observations in each class.
```{r heatmap, fig.asp=1}
print(labels)
print(dim(data))

diff <- distances(data)
```
